---
title: "GRF vs MaxEnt"
author: "Chathura"
date: "November 29, 2016"
output: html_document
---


### Gaussian Random Fields

Using presense data about the species Xantusia Riversiana. 

```{r, echo=FALSE}
url <- "http://www.californiaherps.com/lizards/images/xriversiana104.jpg"
```
<img src="`r url`" height="200px" width="150px">

### Plot the presence on map
```{r, message=FALSE, warning= FALSE}
# load package and data
require(maptools)
require(maps)
require(mapdata)
require(dismo)
require(GRaF)
locs = read.csv(file="./Xantusia_riversiana.csv", header=T, sep=",")
colnames(locs)<-c("ind","species","lat","lon")
plot(c(-119, -110), c(33.5, 38), mar=par("mar"), xlab="longitude", ylab="latitude", xaxt="n", yaxt="n", type="n", main="Xantusia Riversiana Presence")
rect(par("usr")[1],par("usr")[3],par("usr")[2],par("usr")[4], col="lightblue")
map("state", xlim=c(-119, -110), ylim=c(33.5, 38), fill=T, col="cornsilk", add=T)

text(x=-117.5, y=35.5, "California", col="cornsilk3", cex=3)
text(x=-116, y=37.5, "Nevada", col="cornsilk3", cex=3)
text(x=-113, y=34.5, "Arizona", col="cornsilk3", cex=3)
text(x=-113, y=37.75, "Utah", col="cornsilk3", cex=3)

points(locs$lon, locs$lat, col="darkolivegreen4", pch=20, cex=0.5)

axis(1,las=1)
axis(2,las=1)
box()

```

### Correct for sampling bias
```{r,message=FALSE, warning= FALSE}
# create sequences of latitude and longitude values to define the grid
longrid = seq(-119, -110,0.05)
latgrid = seq(33.5, 38,0.05)

# identify points within each grid cell, draw one at random
subs = c()
for(i in 1:(length(longrid)-1)){
  for(j in 1:(length(latgrid)-1)){
    gridsq = subset(locs, lat > latgrid[j] & lat < latgrid[j+1] & lon > longrid[i] & lon < longrid[i+1])     
    if(dim(gridsq)[1]>0){
      subs = rbind(subs, gridsq[sample(1:dim(gridsq)[1],1 ), ])
    }
  }
}
dim(subs) # confirm that you have a smaller dataset than you started with

```

### Generate background data (psuedo-absent) using Raster package.
```{r,message=FALSE, warning= FALSE, collapse=TRUE}
require(raster)

# define circles with a radius of 50 km around the subsampled points
x = circles(subs[,c("lon","lat")], d=50000, lonlat=T)
# draw random points that must fall within the circles in object x
bg = spsample(x@polygons, 1000, type='random', iter=1000)
BClim = getData("worldclim", var="bio", res=2.5, path="data/")
YbrevRange = extent(-119.25,-112.75,33.25,38.25) # define the extent
BClim = crop(BClim, YbrevRange)
writeRaster(BClim, filename="./data/YbrevBC_2.5.grd", overwrite=T)
BClim = brick("data/YbrevBC_2.5.grd")
# this format plots 1 (of 19) variables stored in BClim
plot(BClim, 1, cex=0.5, legend=T, mar=par("mar"), xaxt="n", yaxt="n", main="Annual mean temperature (ºC x 10)")
map("state", xlim=c(-119, -110), ylim=c(33.5, 38), fill=F, col="cornsilk", add=T)
# state names
text(x=-117.5, y=35.5, "California", col=rgb(1,1,1,0.4), cex=3)
text(x=-116, y=37.5, "Nevada", col=rgb(1,1,1,0.4), cex=3)
text(x=-113, y=34.5, "Arizona", col=rgb(1,1,1,0.4), cex=3)
text(x=-113, y=37.75, "Utah", col=rgb(1,1,1,0.4), cex=3)
# plot the presence points
points(locs$lon, locs$lat, pch=20, cex=0.5, col="darkgreen")
# and the pseudo-absence points
points(bg, cex=0.5, col="darkorange3")
# add axes
axis(1,las=1)
axis(2,las=1)
box()
```
# Create dataframe from bioclim and presense/absance.
```{r}
pres<-rep(1,494)
temp1<-data.frame(pres,Ybrev_bc[,3:21])
pres<-rep(0,961)
temp2<-data.frame(pres,bg_bc[,3:21])
df<-rbind(temp1,temp2)
head(df,5)
tail(df,5)
```
### Prior knowldege about the temperature range for  Xantusia Riversiana
Tmax = 36 C
Tmin = 6.6 C

```{r}
covs <- df[1:1455, c("pres","bio1", "bio2", "bio3")]# Not sure these are the best variables.
## 75% of the sample size
smp_size <- floor(0.75 * nrow(covs))
set.seed(123)
train_ind <- sample(seq_len(nrow(covs)), size = smp_size)
train <- covs[train_ind, ]
test <- covs[-train_ind, ]

pa_tr <- train$pres
pa_te <- test$pres
m1 <- graf(pa_tr, train[,2:4])
pred_df<-data.frame(predict(m1,test[,2:4]))
pred_df$posterior.mode
library(pROC)
auc(pa_te, pred_df$posterior.mode)

par(mfrow = c(1, 3))
plot(m1, peak = TRUE)
DIC(m1)


# define the threshold function
thresh <- function(x) ifelse(x$bio1 < 150 | x$bio1 > 250 ,0.2, 0.6)
thresh(covs)
# fit the model, optimising the lengthscale
m3 <- graf(pa_tr, train[,2:4],opt.l = TRUE, prior = thresh)
pred_df<-data.frame(predict(m3,test[,2:4]))
auc(pa_te, pred_df$posterior.mode)
plot(m3, prior = TRUE, main = "threshold prior")
m3
```






```{r}
# pulling bioclim values
Ybrev_bc = extract(BClim, subs[,c("lon","lat")]) # for the subsampled presence points
bg_bc = extract(BClim, bg) # for the pseudo-absence points
Ybrev_bc = data.frame(lon=subs$lon, lat=subs$lat, Ybrev_bc)
bgpoints = bg@coords
colnames(bgpoints) = c("lon","lat")
bg_bc = data.frame(cbind(bgpoints,bg_bc))
length(which(is.na(bg_bc$bio1))) # double-check for missing data
bg_bc = bg_bc[!is.na(bg_bc$bio1), ] # and pull out the missing lines
group_p = kfold(Ybrev_bc, 5) # vector of group assignments splitting the Ybrev_bc into 5 groups
group_a = kfold(bg_bc, 5) # ditto for bg_bc

test = 3

train_p = Ybrev_bc[group_p!=test, c("lon","lat")]
train_a = bg_bc[group_a!=test, c("lon","lat")]
test_p = Ybrev_bc[group_p==test, c("lon","lat")]
test_a = bg_bc[group_a==test, c("lon","lat")]

me = maxent(BClim, p=train_p, a=train_a)
e = evaluate(test_p, test_a, me, BClim)
e


```


```{r}

pred_me = predict(me, BClim) # generate the predictions
# make a nice plot
plot(pred_me, 1, cex=0.5, legend=T, mar=par("mar"), xaxt="n", yaxt="n", main="Predicted presence of the species")
map("state", xlim=c(-119, -110), ylim=c(33.5, 38), fill=F, col="cornsilk", add=T)

# state names
text(x=-117.5, y=35.5, "California", col=rgb(1,1,1,0.6), cex=3)
text(x=-116, y=37.5, "Nevada", col=rgb(1,1,1,0.6), cex=3)
text(x=-113, y=34.5, "Arizona", col=rgb(1,1,1,0.6), cex=3)
text(x=-113, y=37.75, "Utah", col=rgb(1,1,1,0.6), cex=3)

# presence points
points(locs$lon, locs$lat, pch=20, cex=0.5, col="darkgreen")
# pseud-absence points
points(bg, cex=0.5, col="darkorange3")

# add axes
axis(1,las=1)
axis(2,las=1)

# restore the box around the map
box()
```

Incorporating thermal information to species distribution to increase the prediction accuracy.

Intuitively, if some latent structure generalise the environmental variables of a location. We can expect another location with similar laten structure presents favorable conditions for a species to distribute. This notion can be captured in a GRF as follows.

Since we know the species prefered temperature, we can adject the probability of presence based on the temperature of the location. 

