---
title: "prototype_sdm"
author: "Chathura"
date: "November 22, 2016"
output: pdf_document
---

Load data and select reptiles and amphebians for thermal_SDM modelling. The tropical ectothems are used to model. these species have the highly sensitive to temperature so more likely to extinct under global warming.

```{r}
sunday<-read.csv("./Sundayetal_thermallimits.csv",header = T,sep = ",")
tropic_Amphibia <- sunday[sunday$class=="Amphibia" & -23 < sunday$latitude & sunday$latitude < 23,]
tropic_reptilia <-sunday[sunday$class=="Reptilia" & -23 < sunday$latitude & sunday$latitude < 23,]
```

To test the model, select one species and build a correlative SDM. Download from GBIF and extact points in tropical regions.

```{r}
spei<-tropic_reptilia[2,]
print(spei$species)
spe1_gbif<-gbif(genus = as.character(spei$genus),species = as.character(spei$species),geo=T)
spe1<-spe1_gbif[!(is.na(spe1_gbif$lat) |is.na(spe1_gbif$lat) ),]
```
Chamaeleo_dilepis, is a species in africa, so remove other points outside africa and duplicates from same point.
```{r}
spe1_afr<-spe1[-25 < spe1$lon & spe1$lon < 50,]
spe1_afr[1:4,c("class","lat","lon")]
dups <- duplicated(spe1_afr[,c('lon','lat')])
sum(dups)
cham<-spe1_afr[!dups,]
```
cross checking, this may not a big deal.
```{r}
library(sp)
coordinates(cham) <-~lon+lat
crs(cham) <-crs(wrld_simpl)
class(cham)
class(wrld_simpl)
ovr <- over(cham,wrld_simpl)
head(ovr)
cntr <- ovr$NAME
i<-which(is.na(cntr))
i
j <-which(cntr != cham$country)
cbind(cntr,cham$country)[j,]
plot(cham)
plot(wrld_simpl,add=T,border='blue',lwd=2)
points(cham[j,],col='red',pch=20,cex=2)
```
nice plot
```{r}
library(maptools)
data(wrld_simpl)
plot(wrld_simpl, xlim=c(-100,110), ylim=c(-60,30), axes=TRUE,col="light yellow")
box()
points(spe1_afr$lon, spe1_afr$lat, col='orange', pch=20, cex=0.75)
points(spe1_afr$lon, spe1_afr$lat, col='red', cex=0.75)
```

Correcting for sampling bias
```{r}
r <- raster(cham)
res(r) <- 1
r <- extend(r,extent(r) + 1)
cham_sel<-gridSample(cham,r,n=1)
p <- rasterToPolygons(r)
plot(p,border='gray')
points(cham)
points(cham_sel,cex=1,col='red',pch='x')
```

generating absent and background points (psudo-absences) use circles model.

```{r}


files <- list.files(path=paste(system.file(package="dismo"), '/ex',
                                   sep=''), pattern='grd', full.names=TRUE )
# we use the first file to create a RasterLayer
mask <- raster(files[1])
# select 500 random points
# set seed to assure that the examples will always
# have the same random sample.
set.seed(1963)
bg <- randomPoints(mask, 500 )

# set up the plotting area for two maps
par(mfrow=c(1,2))

plot(!is.na(mask), legend=FALSE)
points(bg, cex=0.5)
# now we repeat the sampling, but limit
# the area of sampling using a spatial extent
e <- extent(-80, -53, -39, -22)
bg2 <- randomPoints(mask, 50, ext=e)
plot(!is.na(mask), legend=FALSE)
plot(e, add=TRUE, col='red')
points(bg2, cex=0.5)







cham_sel<-data.frame(cham_sel)
coordinates(cham_sel) <- ~lon+lat
projection(cham_sel) <- CRS('+proj=longlat +datum=WGS84')
x <- circles(cham_sel,d=50000,lonlat=TRUE)
pol <- polygons(x)
samp1 <- spsample(pol,250,type='random',iter=25)
cells <- cellFromXY(mask,samp1)
length(cells)
cells <- unique(cells)

```

Environmental data

```{r}
require(raster) # package to handle raster-formatted spatial data
BClim = getData("worldclim", var="bio", res=2.5, path="data/")
YbrevRange = extent(-25,60,-40,40) # define the extent
BClim = crop(BClim, YbrevRange)
writeRaster(BClim, filename="./data/YbrevBC_2.5.grd", overwrite=T)
BClim <- brick("./data/YbrevBC_2.5.grd")
# this format plots the first (of 19) variables stored in BClim; change the 1 to 2-19 for the others
plot(BClim, 1, cex=0.5, legend=T, mar=par("mar"), xaxt="n", yaxt="n", main="Annual mean temperature (ºC x 10)")
map("state", xlim=c(-119, -113), ylim=c(33.5, 38), fill=F, col="cornsilk", add=T)
```

Example

```{r}
# packages for mapping, and the data for, e.g., state borders
require(maptools)
require(maps)
require(mapdata)
require(dismo)# dismo has the SDM analyses we"ll need
# load the table of latitude and longitude coordinates
locs = read.csv(file="./data/JoTrPresence02202008_dryad.txt", header=T, sep="\t")
# then plot these points to check them ...
data(stateMapEnv) # load the database with the U.S. state borders
# notice we're limiting the extent of the map to focus on the Mojave Desert region
plot(c(-119, -113), c(33.5, 38), mar=par("mar"), xlab="longitude", ylab="latitude", xaxt="n", yaxt="n", type="n", main="Joshua tree presence data")
rect(par("usr")[1],par("usr")[3],par("usr")[2],par("usr")[4], col="lightblue")
map("state", xlim=c(-119, -113), ylim=c(33.5, 38), fill=T, col="cornsilk", add=T)
# add some nice state labels ...
text(x=-117.5, y=35.5, "California", col="cornsilk3", cex=3)
text(x=-116, y=37.5, "Nevada", col="cornsilk3", cex=3)
text(x=-113, y=34.5, "Arizona", col="cornsilk3", cex=3)
text(x=-113, y=37.75, "Utah", col="cornsilk3", cex=3)
# plot the points
points(locs$longitude, locs$latitude, col="darkolivegreen4", pch=20, cex=0.5)
# add some axes
axis(1,las=1)
axis(2,las=1)
box()

```

```{r}

# create sequences of latitude and longitude values to define the grid
longrid = seq(-119,-113,0.05)
latgrid = seq(33.5,38,0.05)

# identify points within each grid cell, draw one at random
subs = c()
for(i in 1:(length(longrid)-1)){
  for(j in 1:(length(latgrid)-1)){
    gridsq = subset(locs, latitude > latgrid[j] & latitude < latgrid[j+1] & longitude > longrid[i] & longitude < longrid[i+1])     
    if(dim(gridsq)[1]>0){
      subs = rbind(subs, gridsq[sample(1:dim(gridsq)[1],1 ), ])
    }
  }
}
dim(subs) # confirm that you have a smaller dataset than you started with
# define circles with a radius of 50 km around the subsampled points
x = circles(subs[,c("longitude","latitude")], d=50000, lonlat=T)
# draw random points that must fall within the circles in object x
bg = spsample(x@polygons, 1000, type='random', iter=1000)

```

```{r}
require(raster) # package to handle raster-formatted spatial data
BClim = getData("worldclim", var="bio", res=2.5, path="data/")
YbrevRange = extent(-119.25,-112.75,33.25,38.25) # define the extent
BClim = crop(BClim, YbrevRange)
writeRaster(BClim, filename="./data/YbrevBC_2.5.grd", overwrite=T)
BClim = brick("data/YbrevBC_2.5.grd")
# this format plots the first (of 19) variables stored in BClim; change the 1 to 2-19 for the others
plot(BClim, 1, cex=0.5, legend=T, mar=par("mar"), xaxt="n", yaxt="n", main="Annual mean temperature (ºC x 10)")
map("state", xlim=c(-119, -113), ylim=c(33.5, 38), fill=F, col="cornsilk", add=T)
# state names
text(x=-117.5, y=35.5, "California", col=rgb(1,1,1,0.4), cex=3)
text(x=-116, y=37.5, "Nevada", col=rgb(1,1,1,0.4), cex=3)
text(x=-113, y=34.5, "Arizona", col=rgb(1,1,1,0.4), cex=3)
text(x=-113, y=37.75, "Utah", col=rgb(1,1,1,0.4), cex=3)
# plot the presence points
points(locs$longitude, locs$latitude, pch=20, cex=0.5, col="darkgreen")
# and the pseudo-absence points
points(bg, cex=0.5, col="darkorange3")
# add axes
axis(1,las=1)
axis(2,las=1)
# restore the box around the map
box()
```

```{r}
# pulling bioclim values
Ybrev_bc = extract(BClim, subs[,c("longitude","latitude")]) # for the subsampled presence points
bg_bc = extract(BClim, bg) # for the pseudo-absence points
Ybrev_bc = data.frame(lon=subs$longitude, lat=subs$latitude, Ybrev_bc)
bgpoints = bg@coords
colnames(bgpoints) = c("lon","lat")
bg_bc = data.frame(cbind(bgpoints,bg_bc))
length(which(is.na(bg_bc$bio1))) # double-check for missing data
bg_bc = bg_bc[!is.na(bg_bc$bio1), ] # and pull out the missing lines
group_p = kfold(Ybrev_bc, 5) # vector of group assignments splitting the Ybrev_bc into 5 groups
group_a = kfold(bg_bc, 5) # ditto for bg_bc

test = 3

train_p = Ybrev_bc[group_p!=test, c("lon","lat")]
train_a = bg_bc[group_a!=test, c("lon","lat")]
test_p = Ybrev_bc[group_p==test, c("lon","lat")]
test_a = bg_bc[group_a==test, c("lon","lat")]

me = maxent(BClim, p=train_p, a=train_a)
train_p = Ybrev_bc[group_p!=test, c("lon","lat")]
train_a = bg_bc[group_a!=test, c("lon","lat")]
test_p = Ybrev_bc[group_p==test, c("lon","lat")]
test_a = bg_bc[group_a==test, c("lon","lat")]

me = maxent(BClim, p=train_p, a=train_a)

e = evaluate(test_p, test_a, me, BClim)
e


```


```{r}

pred_me = predict(me, BClim) # generate the predictions
# make a nice plot
plot(pred_me, 1, cex=0.5, legend=T, mar=par("mar"), xaxt="n", yaxt="n", main="Predicted presence of Joshua trees")
map("state", xlim=c(-119, -113), ylim=c(33.5, 38), fill=F, col="cornsilk", add=T)

# state names
text(x=-117.5, y=35.5, "California", col=rgb(1,1,1,0.6), cex=3)
text(x=-116, y=37.5, "Nevada", col=rgb(1,1,1,0.6), cex=3)
text(x=-113, y=34.5, "Arizona", col=rgb(1,1,1,0.6), cex=3)
text(x=-113, y=37.75, "Utah", col=rgb(1,1,1,0.6), cex=3)

# presence points
points(locs$longitude, locs$latitude, pch=20, cex=0.5, col="darkgreen")
# pseud-absence points
points(bg, cex=0.5, col="darkorange3")

# add axes
axis(1,las=1)
axis(2,las=1)

# restore the box around the map
box()
```



